---
title: "Class 12: Genome informatics"
author: "Isabel Philip (A16855684)"
format: pdf
---

# Section 1. Proportion of G/G in a Population

Downloaded CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39891299-39898900;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read the CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```


```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/ nrow(mxl) *100
```

Now let's look at a different population: GBR population, British in England and Scotland
```{r}
gbr <- mxl <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find Proportion to G|G

```{r}
table(gbr$Genotype..forward.strand.)
```

```{r}
table(gbr$Genotype..forward.strand.)/ nrow(mxl) *100
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population. 

Let's dig into this further: 

## Section 4 Population Analysis 

One sample is obviously not enough to know what is happening in a population. 

You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.


How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes

The sample size of the entire file is 462 (meaning there are 462 samples in total). For the genotype of A/A, there is 108 samples. For the genotype of A/G, there is 233 samples. For the genotype of G/G, there is 121 samples. 


```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

> Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

The boxplot shows that the A/A genotype has higher median expression than the G/G genotype, suggesting a difference in gene regulation. The G allele appears to reduce ORMDL3 expression, while the A allele is associated with higher expression. This suggests that the SNP does affect ORMDL3 expression.
