---
title: "Section 4 - Comparitive Structural Analysis Pt.1"
author: "Isabel Philip (A16855684)"
format: pdf
---

Here we run through s complete "pipeline" of structural analysis that begins with a single seqeunce idenitifer and ends in a PCA analysis. 

```{r}
library(bio3d)
```

## Step 1. Get a sequence 
Step 1: Create a sequence for the protein we are interested in. 
We will take ADK "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

## Step 2. Run a Blast Search 
Run a BLAST search of the PDB for all related seqeunces to our input `aa`

```{r}
#blast <- blast.pdb(aa)
```

```{r}
#hits <- plot(blast)
```

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```


## Step 3. Download Structures 
These are the top hit (18) - i.e. all the structures in the PDB database related to our input sequence. 

```{r}
hits$pdb.id
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

## Step 4. Align and Sequence 

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Step 5. PCA 

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```
> showing there's 3 different clusters/ conformations 

Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the `mktrj()` function for this...

```{r}
mktrj(pc, file ="pca_result.pdb")
```

> Upload file into molstar.org/viewer/ and see through playing video. Easier to view/ understand 


## Back of the Envelope Comparison of the PDB size to UniProt

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot *100
```








