---
title: "Class 10 : Structural Bioinformatics (Pt. 1)"
author: "Isabel Philip (A16855684)"
format: html
---
**Table of contents**

1. PDB database.................................................................. 1
2. Using mol*........................................................................  4
3. Introduction to Bio3D in R...............................................8
4. predicting functional dynamics 12.................................12

**1. PDB database**

The main repository of biomolecular structior data is called the pdb found at: https://www.rcsb.org/
Let’s see what this database contains. PDB analyze >PDB statistics > by Experimental Method and Molecular Type
```{r}
pdbstats <- read.csv("Data Export Summary.csv")
pdbstats
```
commas result in numerical values being categorized as a character

```{r}
pdbstats$X.ray
```
This can be fixed by replacing “,” for nothing “ ” with sub() function:
```{r}
x <- pdbstats$X.ray
x_numeric <- as.numeric( gsub(",", "", x))
x_numeric
```
or I can use the readr package and the read_csv() function

```{r}
library(readr)
pdbstats <- read_csv("Data Export Summary.csv", show_col_types = FALSE)
pdbstats
```

I want to clean the column names so that they are all lower case and don’t have spaces in them

```{r}
colnames(pdbstats)
```

```{r}
library(janitor)
```

```{r}
df <- clean_names(pdbstats)
df
```
```{r}
sum(df$x_ray)
```
```{r}
df$total
```

> Q1: What percentage of structures in the PDB are solved by X-Ray and Electron
Microscopy?

93.58566%
Percent of X-ray structures

```{r}
sum(df$x_ray)/sum(df$total) *100
```
Percent of EM structure
```{r}
sum(df$em)/sum(df$total) *100
```
> Q2: What proportion of structures in the PDB are protein?

86.23852
```{r}
protein_row <- df[df$molecular_type =="Protein (only)",]
protein_row
```

```{r}
sum(protein_row$total) / sum(df$total) * 100
```

> Q3: Type HIV in the PDB website search box on the home page and determine
how many HIV-1 protease structures are in the current PDB?

2,298 structures are currently in the pdb.

**2. Using mol**
The main mol* homepage: https://molstar.org/viewer/ We can input our own PDB files or just give it a PDB database accession code (4 letter PDB code)



